{\huge Supplements for: Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies

### 2024-02-17 Docs/manuscriptRev.tex
revised ms

% recommended! Uncomment the below line and change the path for your computer!
\SweaveOpts{prefix.string=/Users/alina/Documents/git/localadaptclim/Docs/figure_ms/demoFig, eps=FALSE} 
%put your figures in one place! Also, note that here 'figures' is the folder and 'demoFig' is what each 
% figure produced will be titled plus its number or label (e.g., demoFig-nqpbetter.pdf')
% make your captioning look better
\usepackage[small]{caption}
\setlength{\captionmargin}{30pt}
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{10pt}
% optional: muck with spacing
\topmargin -1.5cm        
\oddsidemargin 0.5cm   
\evensidemargin 0.5cm  % same as oddsidemargin but for left-hand pages
\textwidth 15.59cm
\textheight 21.94cm 
% \renewcommand{\baselinestretch}{1.5} % 1.5 lines between lines
\parindent 0pt		  % sets leading space for paragraphs
% optional: cute, fancy headers
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LO]{2024}
\fancyhead[RO]{Manuscript}
\usepackage{lineno} %alinaSep22: this is to number each line (GCB requires it). However, I dont know how not to number the title page, and I also would like the number to appear bigger. Can lizzie fix this please?
\usepackage{setspace}
% Using \doublespacing in the preamble changes the text to double-line spacing
% \doublespacing
\setstretch{1.5} %1.5 spacing
% more optionals! %
% \usepackage[hyphens]{url} % this wraps my URL versus letting it spill across the page, a bad habit LaTeX has
%%% end preambling. %%%

\begin{document}
\SweaveOpts{concordance=TRUE} % For RStudio hiccups
\begin{flushright}
Version dated: \today
\end{flushright}

@@

<<label=source_spring_models, echo=FALSE>>=
#load models and data to report numbers of things
# "C:/Users/alina/Documents/git/localadaptclim"
# setwd("C:/Users/alina/Documents/git/localadaptclim/Docs")
#getwd()
source("../Analyses/source_spring_models_for_sweave.R")
@

Overall, spring events such as budburst and leaf flush were not related to provenance latitude or MAT, neither across continents (latitude: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_mat),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:springfall_latmat}, Table.S\ref{supplement-table:model_spring_lat} \& S\ref{supplement-table:model_spring_mat} in Supporting Information), nor within North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A). Results were similar using other distance metrics in lieu of latitude (see Fig.S\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
% Spring events advanced slightly with provenance latitude in Europe spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes), though uncertainty intervals strongly overlapped zero. --- no longer the case

<<label=source_fall_models, echo=FALSE>>=
#load models and data to report numbers of things
# "C:/Users/alina/Documents/git/localadaptclim"
# setwd("C:/Users/alina/Documents/git/localadaptclim/Docs")Aitken and Bammals (2015
#getwd()
source("../Analyses/source_fall_models_for_sweave.R")
@

In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) were earlier at more northern, cooler MAT sites (that is, they advanced strongly with provenance latitude: \Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_lat)),digits = 2), nsmall=2)} days/degree [\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}], and with decreasing MAT:\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_mat)),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.95), digits = 2)}], Fig.\ref{figure:springfall_latmat}, Table.S\ref{supplement-table:model_fall_lat} \& S\ref{supplement-table:model_fall_mat} in Supporting Information). This relationship, however, was observed mostly in North America where fall events advanced \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C) (Fig.\ref{figure:continental_spptype_effect}A).

@@

Effects of provenance latitude on fall events were similar across angiosperms and gymnosperms (Fig.\ref{figure:continental_spptype_lat}B). Spring events weakly diverged, delaying at a rate of \Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_lat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_lat, 0.95)), digits = 2),nsmall=2)}) days per degree north in angiosperms and advancing \Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in gymnosperms. Fall events advanced \Sexpr{format(round(abs(mean(deciduouspost_fitC_fall_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduouspost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduouspost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in angiosperms and \Sexpr{format(round(abs(mean(evergreenspost_fitC_fall_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in gymnosperms.
Effects of MAT on spring events also weakly diverged (Fig.\ref{figure:continental_spptype_effect}B). Spring events advanced \Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_mat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_mat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_mat, 0.05)), digits = 2),nsmall=2)}) days/$^{\circ}$C as MAT increased in angiosperms and delayed \Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_mat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_mat, 0.95)), digits = 2),nsmall=2)}) days/$^{\circ}$C as MAT increased in gymnosperms. Fall events delayed in warmer locations for both species types, but slightly more so for gymnosperms (\Sexpr{format(round(mean(evergreenspost_fitC_fall_mat),digits = 2),nsmall=2)} days) than angiosperms (\Sexpr{format(round(mean(deciduouspost_fitC_fall_mat),digits = 2),nsmall=2)} days) (Fig.\ref{figure:continental_spptype_effect}B).

While we expected that coarse metrics, such as latitude and MAT, would generally represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months much closer to the events to further test how much climate similarity between provenances and gardens predicts provenance effects (i.e. differential responses observed among plant populations from different geographical origins). For spring events, we considered overlap across March to May. However, results were not qualitatively different than using MAT (See Fig.S\ref{supplement-figure:overlap_scatterplot} in Supporting Information). We observed very weak effects of climate overlap on spring events (\Sexpr{format(round(abs(mean(all_species_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(all_species_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(all_species_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)}] days per one percent increase in climate overlap), nearly identical across angiosperms (\Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)}]) and gymnosperms (\Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)}]). Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing \Sexpr{format(round(abs(mean(evergreenspost_fitC_fall_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.95)), digits = 2),nsmall=2)}] days per one percent decline in climate overlap) (Fig.S\ref{supplement-figure:overlap_posterior} in Supporting Information).

>>>>

% recommended! Uncomment the below line and change the path for your computer!

%put your figures in one place! Also, note that here 'figures' is the folder and 'demoFig' is what each 
% figure produced will be titled plus its number or label (e.g., demoFig-nqpbetter.pdf')
% make your captioning look better
\usepackage[small]{caption}
\setlength{\captionmargin}{30pt}
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{10pt}
% optional: muck with spacing
\topmargin -1.5cm        
\oddsidemargin 0.5cm   
\evensidemargin 0.5cm  % same as oddsidemargin but for left-hand pages
\textwidth 15.59cm
\textheight 21.94cm 
% \renewcommand{\baselinestretch}{1.5} % 1.5 lines between lines
\parindent 0pt		  % sets leading space for paragraphs
% optional: cute, fancy headers
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LO]{2024}
\fancyhead[RO]{Manuscript}
\usepackage{lineno} %alinaSep22: this is to number each line (GCB requires it). However, I dont know how not to number the title page, and I also would like the number to appear bigger. Can lizzie fix this please?
\usepackage{setspace}
% Using \doublespacing in the preamble changes the text to double-line spacing
% \doublespacing
\setstretch{1.5} %1.5 spacing
% more optionals! %
% \usepackage[hyphens]{url} % this wraps my URL versus letting it spill across the page, a bad habit LaTeX has
%%% end preambling. %%%

\begin{document}
\input{/Users/alina/Documents/git/localadaptclim/Docs/figure_ms/demoFig-concordance} % For RStudio hiccups
\begin{flushright}
Version dated: \today
\end{flushright}

@@

Overall, spring events such as budburst and leaf flush were not related to provenance latitude or MAT, neither across continents (latitude: 0.10 days/degree [-0.05 - 0.25]; MAT: -0.11 days/$^{\circ}$C [-0.34 - 0.12]) (Fig.\ref{figure:springfall_latmat}, Table.S\ref{supplement-table:model_spring_lat} \& S\ref{supplement-table:model_spring_mat} in Supporting Information), nor within North America (latitude: 0.10 days/degree [-0.06 - 0.26]; MAT: -0.09 days/$^{\circ}$C [-0.36 - 0.18]) or Europe (latitude: 0.10 days/degree [-0.23 - 0.42]; MAT: -0.16 days/$^{\circ}$C [-0.55 - 0.23]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A). Results were similar using other distance metrics in lieu of latitude (see Fig.S\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
% Spring events advanced slightly with provenance latitude in Europe spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes), though uncertainty intervals strongly overlapped zero. --- no longer the case

In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) were earlier at more northern, cooler MAT sites (that is, they advanced strongly with provenance latitude: 3.16 days/degree [2.87-3.45], and with decreasing MAT:4.78 days/$^{\circ}$C [4.1 - 5.4], Fig.\ref{figure:springfall_latmat}, Table.S\ref{supplement-table:model_fall_lat} \& S\ref{supplement-table:model_fall_mat} in Supporting Information). This relationship, however, was observed mostly in North America where fall events advanced 4.24 (3.95 - 4.53) days per degree northward, or 6.41 days (5.78 - 7.04) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of 0.47 (0.21 - 1.17) days per degree northward, or 0.70 days (1.04 - 2.42) per degree decline in MAT ($^{\circ}$C) (Fig.\ref{figure:continental_spptype_effect}A).

@@

Effects of provenance latitude on fall events were similar across angiosperms and gymnosperms (Fig.\ref{figure:continental_spptype_lat}B). Spring events weakly diverged, delaying at a rate of 0.37 (0.15 - 0.59) days per degree north in angiosperms and advancing 0.23 (0.00 - 0.46) days per degree north in gymnosperms. Fall events advanced 3.18 (2.76 - 3.62) days per degree north in angiosperms and 3.14 (2.81-3.47) days per degree north in gymnosperms.
Effects of MAT on spring events also weakly diverged (Fig.\ref{figure:continental_spptype_effect}B). Spring events advanced 0.82 (0.54 - 1.11) days/$^{\circ}$C as MAT increased in angiosperms and delayed 0.76 (0.37 - 1.14) days/$^{\circ}$C as MAT increased in gymnosperms. Fall events delayed in warmer locations for both species types, but slightly more so for gymnosperms (6.23 days) than angiosperms (3.69 days) (Fig.\ref{figure:continental_spptype_effect}B).

While we expected that coarse metrics, such as latitude and MAT, would generally represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months much closer to the events to further test how much climate similarity between provenances and gardens predicts provenance effects (i.e. differential responses observed among plant populations from different geographical origins). For spring events, we considered overlap across March to May. However, results were not qualitatively different than using MAT (See Fig.S\ref{supplement-figure:overlap_scatterplot} in Supporting Information). We observed very weak effects of climate overlap on spring events (0.01 [0.02 - 0.03] days per one percent increase in climate overlap), nearly identical across angiosperms (0.02 [0.00 - 0.05]) and gymnosperms (0.04 [0.00 - 0.09]). Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing 0.72 [0.51 - 0.92] days per one percent decline in climate overlap) (Fig.S\ref{supplement-figure:overlap_posterior} in Supporting Information).
